---
title: "Problem Set 4 — my solution"
author: "Zhengting (Johnathan) He"
date: "2021/10/7"
output: html_document
---


```{r "setup", include = FALSE}
require("knitr")
opts_knit$set(root.dir = "D:/OneDrive - Johns Hopkins/Course/140.621.81 - Statistical Methods in Public Health I/Problem set/jhsphbiostat621-assignment/Problem Set 4")
```

```{css}
blockquote {
    padding: 10px 20px;
    margin: 0 0 20px;
    font-size: 14px;
    border-left: 5px solid #eee;
}
```

```{r, message = FALSE}
# Set up
setwd("D:/OneDrive - Johns Hopkins/Course/140.621.81 - Statistical Methods in Public Health I/Problem set/jhsphbiostat621-assignment/Problem Set 4")
```


# Problem Set 4: Confidence Intervals and Hypothesis Tests (with R)


## Problem 1. The Possible Association between Particulate Air Pollution and Daily Mortality in Baltimore


0. Load the tidyverse with: `library(tidyverse)` and import the data set and name it
`balt621`. Also, don’t forget to open a script file in order to save your work.

```{r, message = FALSE}
require(tidyverse)
balt621 <- read_csv("./data/balt621.csv")
```


i. Stratify the daily mortality data into 4 groups by season using the season variable.
Calculate the mean PM10 level and mean daily mortality by season.

```{r}
balt621 %>%
    group_by(season) %>%
    summarize(n_pm10 = sum(!is.na(pm10)),
              mean_pm10 = mean(pm10, na.rm=TRUE),
              n_mortality = sum(!is.na(death)),
              mean_mortality = mean(death)) %>%
    kable()
```


ii. For each of the four seasonal strata, stratify the days into 5 pollution (pm10) strata.

```{r, results = 'asis'}
pm10.season <- list() # Empty list for each seasonal strata
pm10.season.table <- list()

for(seas in c("Autumn", "Spring", "Summer", "Winter")) {
    pm10.season[[seas]] <-
        balt621 %>%
        filter(season == seas) %>%
        mutate(pm_group = cut(pm10,
                              breaks =  quantile(.[, "pm10"],
                                                 c(0, .2, .4, .6, .8, 1),
                                                 na.rm = TRUE),
                              labels = 1:5))
    
    pm10.season.table[[seas]] <- table(pm10.season[[seas]]$pm_group)

}

bind_rows("Autumn" = pm10.season.table[["Autumn"]],
          "Spring" = pm10.season.table[["Spring"]],
          "Summer" = pm10.season.table[["Summer"]],
          "Winter" = pm10.season.table[["Summer"]],
          .id = "Season.Strata") %>%
    kable()
    
```


iii. Separately for your four seasonal strata, calculate the mean mortality in the lowest and
highest of the five pollution strata.

```{r, results = 'asis'}
pm10.season.summary <- list()
for(seas in c("Autumn", "Spring", "Summer", "Winter")) {
   pm10.season.summary[[paste0(seas, "1")]] <-
   pm10.season[[seas]] %>%
       filter(pm_group == 1) %>%
       summarize(mean = mean(death),
                 sd = sd(death),
                 n = n())
   
   pm10.season.summary[[paste0(seas, "5")]] <-
   pm10.season[[seas]] %>%
       filter(pm_group == 5) %>%
       summarize(mean = mean(death),
                 sd = sd(death),
                 n = n())
}

bind_rows("Autumn:Low" = pm10.season.summary[["Autumn1"]],
          "Autumn:High" = pm10.season.summary[["Autumn5"]],
          "Spring:Low" = pm10.season.summary[["Spring1"]],
          "Spring:High" = pm10.season.summary[["Spring5"]],
          "Summer:Low" = pm10.season.summary[["Summer1"]],
          "Summer:High" = pm10.season.summary[["Summer5"]],
          "Winter:Low" = pm10.season.summary[["Winter1"]],
          "Winter:High" = pm10.season.summary[["Winter5"]],
          .id = "Group") %>%
    separate(Group, c("Season", "Strata"), ":") %>%
    select(Strata:n) %>%
    kable() %>%
    kableExtra::pack_rows(index = c("Autumn" = 2, "Spring" = 2, "Summer" = 2, "Winter" = 2)) %>%
    kableExtra::kable_styling()
```


iv. Within each seasonal stratum, test the null hypothesis that the mean mortality is the same
regardless of the pollution level by comparing the mean mortality in the lowest and
highest pollution strata. *Given the large sample sizes, there is no need to pool the
variances.*

```{r}
pm.season.15 <- list()

for(seas in c("Autumn", "Spring", "Summer", "Winter")) {
    print(paste("# T-test of mean mortality in lowest and highest pollution strata in",
                seas, sep = " "))
    
    pm.season.15[[seas]] <-
        pm10.season[[seas]] %>%
        filter(pm_group==1 | pm_group==5)
    
    print(t.test(death ~ pm_group, data = pm.season.15[[seas]], var.equal = FALSE))
}
```


v. For each seasonal stratum, calculate by hand a 95% confidence interval for the true mean
mortality difference between the lowest and highest pollution days. 

<blockquote>
1. Autumn:

$$
\bar{X_{1}} - \bar{X_{2}} \pm t_{0.05/2, df=221} * s_{\bar{X_{1}} - \bar{X_{2}}}
      = 19.82 - 20.53 \pm 1.97 * \sqrt{\frac{5.43^{2}}{112} + \frac{5.38^{2}}{111}}
      = [-2.14, 0.72]
$$

2. Spring:

$$
\bar{X_{1}} - \bar{X_{2}} \pm t_{0.05/2, df=217} * s_{\bar{X_{1}} - \bar{X_{2}}}
      = 18.55 - 18.99 \pm 1.97 * \sqrt{\frac{4.19^{2}}{112} + \frac{5.01^{2}}{113}}
      = [-1.65, 0.77]
$$

3. Summer:

$$
\bar{X_{1}} - \bar{X_{2}} \pm t_{0.05/2, df=207} * s_{\bar{X_{1}} - \bar{X_{2}}}
      = 17.63 - 19.93 \pm 1.97 * \sqrt{\frac{4.15^{2}}{114} + \frac{5.68^{2}}{114}}
      = [-3.60, -1.00]
$$

4. Winter:

$$
\bar{X_{1}} - \bar{X_{2}} \pm t_{0.05/2, df=218} * s_{\bar{X_{1}} - \bar{X_{2}}}
      = 22.24 - 21.35 \pm 1.97 * \sqrt{\frac{5.30^{2}}{110} + \frac{5.35^{2}}{110}}
      = [-0.52, 2.30]
$$
</blockquote>


vi. In a table, for each of the four seasonal strata, show the mean mortality on low pollution 
days and its interval estimate, the mean mortality on high pollution days and its interval 
estimate, and the difference in mean mortality with interval estimate for the difference in 
mean mortality between low and high pollution days.

<blockquote>
|Season|Mortality on low pollution days </br> mean[95%CI]|Mortality on high pollution days </br> mean[95%CI]|Difference in mean mortality </br> mean[95%CI]|
|:--:|:--:|:--:|:--:|
|Autumn|19.82 [18.81, 20.83]|20.53 [19.51, 21.54]|-0.71 [-2.14, 0.72]|
|Spring|18.55 [17.77, 19.34]|18.99 [18.06, 19.92]|-0.44 [-1.65, 0.77]|
|Summer|17.63 [16.86, 18.40]|19.93 [18.88, 20.98]|-2.30 [-3.60, -1.00]|
|Winter|22.24 [21.24, 23.24]|21.35 [20.34, 22.37]|0.88 [-0.52, 2.30]
</blockquote>


vii. Summarize your findings in a few sentences about whether mortality is associated with 
pollution (PM10) when controlling for seasonality and whether season modifies this 
association. Be quantitative citing mortality differences with their CIs.

<blockquote>
When setting the statistical significant level as \alpha = 0.05, as shown in the results above, only within the summer subgroup, confidence interval of the difference in mean mortality between low and high pollution days did not cover the null value 0, and p-value < 0.05 with t-score within rejection region; thus can reject the null hypothesis. However, within autumn, spring, or winter subgroup, confidence interval of the difference in mean mortality between low and high pollution days all cover the null value 0, and p-value > 0.05 with t-score out of the rejection region; thus cannot reject the null hypothesis. These results indicate that when controlling for seasonality, **mortality is associated with pollution (PM10) *only* within the summer subgroup**, and not associated with pollution (PM10) within autumn, spring, or winter subgroup. Since the association between mortality and pollution (PM10) varies, **seasonality is an effect modifier for this association**.
</blockquote>

